Capturing the shape of our experiences and how we remember them

### 2018-08-17 paper/main.tex
saving progress
more paper reorganization

\title{Capturing the shape of our experiences and how we remember them}
% \title{The fundamental shape of naturalistic experiences is preserved in episodic memory}
% \title{Memory dynamics for naturalistic experiences}
\author{Andrew C. Heusser \& Jeremy R. Manning}
% \affiliation{Dartmouth College}

@@

\section{Abstract}
{
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that they unfold in a structured and predictable manner, governed by the laws of the physical systems that produce them. However, this temporal structure is typically absent (or severly impoverished) in traditional trial-based memory experiments. As a consequence, our understanding of how the inherent temporal structure of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the stimulus and participants' recalls. We found that despite large differences in the amount of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by most of the participants. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.

% The human memory system is adept at cataloging the rich dynamics of ongoing experience. However, traditional trial-based memory experiments cannot capture these dynamics, and therefore cannot be used to study them.  By constraining participants' experiences in an experiment to occur in temporally discrete trials (and often arranged in a randomized order) the temporal, contextual, and emotional structure of those in-lab experiences necessarily differ from the naturalistic experiences we encounter in everyday life.  Here we investigate how people verbally recall continuous videos by characterizing and relating the thematic dynamics, or ``trajectories,'' of the stimulus and participants' recalls.  Unlike trial-based studies of memory wherein participants attempt to recall the precise stimuli they encounter, naturalistic recall entails capturing the fundamental geometric components of the stimulus topic trajectory.  The precise words participants use to describe the stimulus, and the level of detail and number of distinct events they recall, vary considerably across participants.  Nevertheless, the majority of participants' recall narratives captured the fundamental ``shape'' of the original stimulus.  These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a binary operation: either an item is recalled or it isn't. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' an experience or a feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., telling a story to a friend about a recent vacation) is much more nuanced. Real-world recall is continuous, rather than binary.  The specific words used to describe an experience have very little bearing on whether the experience is considered to have been ``remembered.''  Further, one might remember the gist of an experience but forget (or neglect to recount) particular details.  Or different people who share an experience might recount the experience with a similar level of detail, but the specific details that were remembered might vary across people.  Which aspects of those recollections should be considered fundamental, and which are extraneous to the main story?

Human memory studies typically assess memory for the ``items'' within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e. how the contents relate/change over time).  Naturalistic experiences are typically autocorrelated in space and time on short timescales: the contents of an experience are often similar from moment to moment and change gradually. For example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall specific contents of an experience is one way to assess memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

% I realize this sticks out and needs to be better integrate into the narrative, but not sure how!
Despite the continuous and (often) gradually changing nature of naturalistic experiences, prior research suggests that our memories for those same experiences may be more descrete and organized around particular ``events'' \citep{Radv12, RadvZack11, BrunEtal18}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience (i.e. ``event boundaries'') may lead to discontinuous in memory \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

In this study, we analyzed an open dataset in which participants view and then verbally recount an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the temporal structure present in naturalistic experiences, we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. We propose that the temporal structure of an experience gives it a unique geometric ``shape'' or ``trajectory'', defined by how the contents of the experience relate and change over time. We hypothesized that when recounting naturalistic experiences, the precise details that are recapitulated and language used may differ across people, but the trajectory of the stimulus is generally preserved.

@@

\subsubsection{A model to capture the temporal and semantic structure of naturalistic stimuli}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.\ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). We then transformed the text descriptions using the (same) topic model, resulting in a scenes (1976) by topics (100) matrix, where each row of the matrix represents a probabalistic mixture of topics discovered in that scene (See Fig.~\ref{fig:schematic} for example topic vectors). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic `state' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure. Another noteworthy characteristic is that there is very little correlation between blocks (i.e. the off-diagonal values are small), suggesting the representations of each block are unique and highly descriminable.

\subsubsection{Modeling verbal recall}
After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. The result was a sentences (range: 68 to 294, mean: 131.94) by topics (100) matrix for each participant, where each row represents a probabalistic mixture of topics for a given window of sentences during recall. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Like the video model, topic vectors were sparse and changed gradually.  Note that the topics were derived solely from the text descriptions of the video, and so the topic models estimated for recall are dependent on the topics present in the video. This approach effectively projects the participants' verbal recall into the video's ``topic space'', allowing us to directly compare them to the video model.

Next, we investigated the temporal structure of the recall matrices. For each participant, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix from the recall models. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d). Like the video model, each participants' recall correlation matrix exhibited a strong block-diagonal structure (Supp Fig. 1). Notably, this suggests that participants recounted the video in discriminable segments, potentially related to the recall of specific events (or scenes) from the video.

\subsubsection{Segmenting the video and recall models into `events'}
As described above, a striking feature of the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in the contents of the video (i.e. ``events''), as well as the subsequent language used to recount each event of the video. If true, we reasoned that it would be possible to identify the specific event a participant is describing by comparing the topic vector for a particular recall ``event'' with the topic vector for each video event. In other words, the video event that is most similar (i.e. correlated) to a particular recall event is the event that the participant is most likely describing.

To test this idea, first we segmented the video and recall models in $k$ events (i.e. states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the model and a representative participant (\#13) are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model by'' averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (see Fig.~\ref{fig:model}f for example). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Interestingly, our algorithm suggests that the example participant recalled most of the video events in order.

@@

\subsubsection{Characterizing naturalistic memory with traditional list-learning analyses}
Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. In our first set of analyses, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings.

First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where subjects with a greater number of recall events also had better overall memory performance (Pearson's $r$(16)=.67, $p$=.003). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). We found that participants tended to initiate their recall sequences with the first few events (Fig.~\ref{fig:list-learning}a), which is qualitatively very similar to previously published list learning experiments (REF?). Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. The result suggests a strong bias to transition sequentially events in the forward direction (Fig.~\ref{fig:list-learning}b). Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Fig.~\ref{fig:list-learning}c). We found that there was substantial variability in memory for the different events (STAT).

We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. We found that participants who clustered in time also recalled a greater number of events (Pearson's $r$(16)=.62, $p$=.007). Next, we assessed semantic clustering. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r$(16)=.55, $p$=.02).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.

\subsubsection{The shape of a naturalistic experience is preserved in recall}
The analyses described in the previous sections are useful in that they allow us to quantify memory dynamics for naturalistic stimuli in a way that is comparable to a vast literature on trial-based list-learning experiments. Furthermore, we introduce novel measures (precision and distinctiveness) which provide additional details regarding the veracity of naturalistic memories. However, these methods fail to capture the rich temporal structure present in naturalistic stimuli and associated recall of those experiences. Thus, our next set of analyses test whether the ``shape'' of an experience is preserved in memory. We hypothesized that despite substantial individual variability in the precision and amount of content recalled across by participants, the overall shape of the episode would be recapitulated by the majority of participants.

@@

To visualize the relationship between the video and recall event models, we embedded them into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity in ``topic space'' (Fig.~\ref{fig:trajectory}a).   We observed that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p$<$.001, corrected at p$<$.05 using permutation procedure described in Methods).Thus participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

Notably, many of the participants' recall trajectories appear to recover the overall shape of the video stimulus. To quantify this similarity, we measured the average euclidean distance between each each segment of the video model (e.g. the line formed by successive video events) and the closest recall event for each participant.  Inuitively, if the participant approximately remembered every video event, the measure would approach 0. The more missing recall events, the higher the value. We found that this shape measure was highly correlated to the number of events remembered across participants (Pearson's $r$(16)=-.87, p$<$.001). While these measures are highly correlated (as might be expected), there are cases where the measures dissociate.  For example participant X and X recalled exactly the same number of events, but the shape measure is drastically different between them.  Conversely, participant X and X had the same shape measure, but there was a large difference in the number of events recalled. Thus, measuring similarity in shape between the video and recall models provides unique information about participant's recall behavior that is not captured by classic measures of memory accuracy.

\subsubsection{Measuring the quality of naturalistic memories}
Representing the video and verbal recall as events in ``topic space'' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than what was previously possible. To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered. Then, we computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p$<.05; participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to hand annotated memory performance across participants (Pearson's $r$(16)=-.57, $p$=.016). Thus, a closer match between the video and recall event models was related to better recall performance.

Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled). This resulted in a single number for each recalled event indexing how similar the recall event was to its matching movie event (i.e the ``precision'' of the recall). We then averaged the correlations within participant. In line with our prediction, there was a strong correlation between hand annotated memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's $r$(16)=.74, p=.0006). Next, we considered the distinctiveness of each recall event. That is, how uniquely a recall event matched a given video event compared to all other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way). To this end, we computed a `distinctiveness' score for each participant (i.e. 1 - the correlation between a recall event and all non-matching video events).  Then, we averaged this measure over recall events within participant.  We found that participants with higher hand annotated memory performance also had higher distinctiveness scores (Pearson's $r$(16)=.8, p=.0001).

Lastly, we tested whether participants with better memory performance were also more likely to remember the events in order.  For each participant, we computed the Spearman rank correlation between the order of events that the participant recalled and the actual order of events (filtering events that were actually recalled).  We found that participants who recalled more events also recalled more of them in order (Pearson's $r$(16)=.5, p=.04). In summary, we found that better memory performance was associated with more precise, distinctive and ordered recalls.

@@

@@

To conclude, let's revisit the question of what it means to remember something. We assert that episodic memories are not veridical snapshots from the past, and so we should stop treating (and modelling) them as such.  Rather, episodic memories are better described as echoes of a prior experience that are shaped by other past experiences as well as the current context. Furthermore

@@

% Supplemental materials

\begin{suppfigure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/supp1_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e. topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{suppfigure}

\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp2_corrmats.pdf}
\caption{\small \textbf{Recall model correlation matrices and event segmentation fits.} Each participant's timepoint-by-timepoint recall correlation matrix.  The yellow boxes represent ``events'' identified by a hidden Markov model.}
\label{fig:corrmats}
\end{suppfigure}

>>>>

\title{Capturing the shape of our experiences and how we remember them}
\author{Andrew C. Heusser \& Jeremy R. Manning}
% \affiliation{Dartmouth College}

@@

\section{Abstract}
{
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that they unfold in a structured and predictable manner, governed by the laws of the physical systems that produce them. However, this temporal structure is typically absent (or severly impoverished) in traditional trial-based memory experiments. As a consequence, our understanding of how the temporal structure (i.e. the shape) of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the stimulus and participants' recalls. Using a novel approach based on topic modelling, we found that despite substantial variability in the quantity of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by most of the participants. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a binary operation: either an item is recalled or it isn't. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' an experience or a feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., telling a story to a friend about a movie you recently saw) is much more nuanced. Real-world recall is continuous, rather than binary.  The specific words used to describe an experience have very little bearing on whether the experience is considered to have been ``remembered.''  Further, one might remember the gist of an experience but forget (or neglect to recount) particular details.  Or different people who share an experience might recount the experience with a similar level of detail, but the specific details that were remembered might vary across people.  Which aspects of those recollections should be considered fundamental, and which are extraneous to the main story?

Human memory studies typically assess memory for the ``items'' within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e. how the contents relate/change over time).  Naturalistic experiences are (typically) autocorrelated in space and time on short timescales: the contents of an experience are often similar from moment to moment and change gradually. For example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall specific contents of an experience is one way to assess memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

Despite the continuous and (often) gradually changing nature of naturalistic experiences, prior research suggests that our memories for those same experiences may be more descrete and organized around particular ``events'' \citep{Radv12, RadvZack11, BrunEtal18}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience (i.e. ``event boundaries'') may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

In this study, we analyzed an open dataset in which participants view and then verbally recount an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the flow of the video's content (i.e. the storyline), we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. We propose that the temporal evolution of an experience's content gives it a unique geometric ``shape'' or ``trajectory'', defined by how the contents of the experience relate and change over time. We hypothesized that when recounting naturalistic experiences, the precise details remembered and language used may differ across people, but the shape of the stimulus would be preserved.

@@

\subsubsection{A model to capture the temporal and semantic structure of naturalistic stimuli}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.\ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic `state' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable.

\subsubsection{Modeling verbal recall}
After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d). Like the video model, each participants' recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats})). Notably, this suggests that participants recounted the video in discriminable segments, potentially related to the recall of specific events (or scenes) from the video.

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/supp1_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e. topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\subsubsection{Segmenting the video and recall models into `events'}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in "theme" of the video/recall (i.e. ``events''). If true, we reasoned that specific events described by a participant could be ``matched'' to specific video events by computing the similarity between their topic vectors. In other words, the video event that is most similar (i.e. correlated) to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e. states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the model and a representative participant (\#13) are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (see Fig.~\ref{fig:model}f for example). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Interestingly, our algorithm suggests that the example participant recalled most of the video events in order.

@@

\subsubsection{The shape of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization. However, none of these approaches capture the rich autocorrelated structure inherent in naturalistic stimuli and associated recall of those experiences. Using our topic-based approach, in the next set of analyses we test the hypothesis that naturalistic recall entails recapitulating the trajectory of a stimulus.

@@

To visualize the relationship between the video and recall event models, we embedded them into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity in ``topic space'' (Fig.~\ref{fig:trajectory}a).   We observed that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p$<$.001, corrected at p$<$.05 using permutation procedure described in Methods). Thus participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

% NOTE: Not sure if this part (below) is necessary, but if it is, then I think we need to come up with a null test to say that participants fits are better than chance

% Notably, many of the participants' recall trajectories appear to recover the shape of the video stimulus. To quantify this similarity, we measured the average euclidean distance between each each segment of the video model (e.g. the line formed by successive video events) and the closest recall event for each participant.  Inuitively, if the participant approximately remembered every video event, the measure would approach 0. The more missing recall events, the higher the value. We found that this shape measure was highly correlated to the number of events remembered across participants (Pearson's $r$(16)=-.87, p$<$.001). While these measures are highly correlated (as might be expected), there are cases where the measures dissociate.  For example participant X and X recalled exactly the same number of events, but the shape measure is drastically different between them.  Conversely, participant X and X had the same shape measure, but there was a large difference in the number of events recalled. Thus, measuring similarity in shape between the video and recall models provides unique information about participant's recall behavior that is not captured by classic measures of memory accuracy.

@@

\subsubsection{Additional measures of naturalistic memory}
Representing the video and verbal recall as events in ``topic space'' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than what was previously possible. To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered. Then, we computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p$<.05; participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to hand annotated memory performance across participants (Pearson's $r$(16)=-.57, $p$=.016). Thus, a closer match between the video and recall event models was related to better recall performance.

Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled). This resulted in a single number for each recalled event indexing how similar the recall event was to its matching movie event (i.e the ``precision'' of the recall). We then averaged the correlations within participant. In line with our prediction, there was a strong correlation between hand annotated memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's $r$(16)=.74, p=.0006). Next, we considered the distinctiveness of each recall event. That is, how uniquely a recall event matched a given video event compared to all other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way). To this end, we computed a `distinctiveness' score for each participant (i.e. 1 - the correlation between a recall event and all non-matching video events).  Then, we averaged this measure over recall events within participant.  We found that participants with higher hand annotated memory performance also had higher distinctiveness scores (Pearson's $r$(16)=.8, $p$=.0001).

Lastly, we tested whether participants with better memory performance were also more likely to remember the events in order.  For each participant, we computed the Spearman rank correlation between the order of events that the participant recalled and the actual order of events (filtering events that were actually recalled).  We found that participants who recalled more events also recalled more of them in order (Pearson's $r$(16)=.5, $p$=.04). In summary, we found that better memory performance was associated with more precise, distinctive and ordered recalls.

Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. In our first set of analyses, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings. First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where subjects with a greater number of recall events also had better overall memory performance (Pearson's $r$(16)=.67, $p$=.003). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). We found that participants tended to initiate their recall sequences with the first few events (Supp. Fig.~\ref{fig:list-learning}a), which is qualitatively very similar to previously published list learning experiments (REF?). Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. The result suggests a strong bias to transition sequentially events in the forward direction (Supp. Fig.~\ref{fig:list-learning}b). Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Supp. Fig.~\ref{fig:list-learning}c). We found that there was substantial variability in memory for the different events (STAT). We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. We found that participants who clustered in time also recalled a greater number of events (Pearson's $r$(16)=.62, $p$=.007). Next, we assessed semantic clustering. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r$(16)=.55, $p$=.02).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.

@@

To conclude, let's revisit the question of what it means to remember something. We assert that episodic memories are not veridical snapshots of the past, and so we should stop treating (and modelling) them as such.  Rather, episodic memories are better described as echoes of a prior experience that are shaped by other past experiences as well as the current context.

@@

% Supplemental materials
\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp2_corrmats.pdf}
\caption{\small \textbf{Recall model correlation matrices and event segmentation fits.} Each participant's timepoint-by-timepoint recall correlation matrix.  The yellow boxes represent ``events'' identified by a hidden Markov model.}
\label{fig:corrmats}
\end{suppfigure}
